LOSD: A Linked Open Social Dataset for Scientific Research

### 2019-07-02 mdpi/template.tex
acrescentado sobre virtual ethnography
finalizada revisao do artigo

%=================================================================
% Full title of the paper (Capitalized)
\Title{LOSD: A Linked Open Social Dataset for Scientific Benchmarking}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
The fields of social network analysis and complex networks
are widely researched.
In fact, a myriad of results have been reported which are based in
diverse data, most often not accessible to researchers other than the publishing authors.
In order to provide the scientific community a common repertoire,
this work presents an open dataset with diverse provenance of 
human social online networking.
Current data was obtained in 2013-16 from well-known online social networking platforms:
Facebook, Twitter, IRC, Email.
Data from ParticipaBR, AA, and Cidade Democrática are also supplied as samples of less known instances.
The linked data representation was adopted in order to
comply with current best practices, homogenize access, facilitate discovery
and analyzes which integrate third party and provided data.
Futhermore, the method for obtaing and making the data available is potentially novel,
and may be of use for other research groups.
This document presents a description, usage remarks and overall statistics of the dataset,
which should favor subsequent work in complex and social networks and enhancements to the dataset.
}

@@

\begin{document}
\section{Summary}
The research on human social and complex networks often
rely on data that express real phenomena~\cite{c1,c2,c3}.
Such data is regularly not made publicly available,
which hinders validation of the results
and does not favor subsequent work.
Furthermore, there is a lack of open datasets for benchmarking results
related to complex and social networks,
yielding diverse results from poorly related sources.
Social networks linked datasets are currently most often restricted to a specific platform or express statistics of social networks, as described in Section~\ref{srel}.
This work presents 
a dataset named LOSD (Linked Open Social Data)
that has been achieved by linking social data derived from diverse
provenance, namely: Facebook, Twitter, IRC, Email groups, ParticipaBR, Cidade Democrática, and AA.
Such data may be of use as a common and public repertoire for scientific
research involving complex networks, specially if social networks and textual content are pinpointed.
In fact, the data has already been used in scientific research~\cite{stab,thesis} (CNPq grant 140860/2013-4),
and should be further used e.g. in researching complex networks and data visualization (FAPESP grant 2017/05838-3)
which relies on networks enriched by textual data and on dynamic (or time-evolving) networks.
Data collection was performed by a potentially innovative method which places the researcher
at the core of data collection, and keeps the resources (e.g. code, data, documents) open,
in order to ammeliorate ethical issues involved in studying human social systems.

This paper is organized as follows. Next subsection briefly discusses related work.
Section~\ref{materials} describes the data currently available.
Section~\ref{methods} describes the methods used for data gathering, translation into RDF linked data, and the synthesis of auxiliary structures.
Section~\ref{usage} holds directions for using the data.

\subsection{Related work}\label{srel}
Most available datasets on human social networks rely only on one source (e.g.~\cite{nat,fb1}),
on specialized communities such as related to research or movies (e.g.~\cite{s1},
or present only statistics about social networks, and not the data that yields the networks (e.g.~\cite{st1}.
Linked human social network data is scarcely found in the scientific community.
In~\cite{foaf1}, the authors presented a set of tools to obtain linked social data available in the public data cloud, and made available a dataset from the Advogato platform,
dedicated to free software developers linked by trust relatioships.
There are various datasets which makes available data obtained from Twitter (e.g.~\cite{tw1,tw2}),
most often with a specific focus (users, retweets, hashtags, etc)
and not presented to the scientific community by means of a research article.
The datasets which are most closely related to LOSD are KONECT~\cite{konect},
SNAP Datasets~\cite{snapnets}, the Network Repository~\cite{nr}, and ICON~\cite{icon}.
None of them are (RDF) linked data but express the networks in formats specialized for networks.
In total, these datasets make available thousands of social networks, from diverse provenance and of various types (e.g. simple, directed, weighted, bipartite, temporal networks).
Compared to LOSD, KONECT, SNAP, Network Repository, and ICON are complementary, providing a further diversity of networks, but only as simplified data, i.e. in specialized formats, most often expressed as columns of node ids that are connected and metadata (such as timestaps for temporal networks).
In summary, no linked human social network dataset presented to the
scientific community containts the diverse provenance used in LOSD,
or provides such a wide range of metadata,
potentially because researchers did not have the means to acquire and
share the data as described in Section~\ref{acq},
and because linked data standards are (still) not widely adopted in 
every research domain.

\section{Data Description}\label{materials}
\subsection{Data outline}\label{outline}
% scripts/percolationUse2.py
The database consists of 31,996,740 triples, 3,030,433 network links yield by main interactions or relations, 350,284 participants and 245,194,377 characters. Among all snapshots, 63 are ego snapshots, 54 are group snapshots; 50 have interaction edges, 89 have friendship edges; 43 have text content from messages.
Section~\ref{queries} holds the queries used to obtain such data.
This section describes de data in LOSD in very brief terms.
The Supplementary document is provided for further information about the dataset.

\noindent Data was gathered from:
\begin{itemize}
    \item public APIs (Twitter, Email);
    \item public logs (IRC and AA);
    \item Netvizz software~\cite{netvizz} and subsequent donation by users (Facebook);
    \item donation by system administrators (AA, ParticipaBR, Cidade Democr\'atica).
\end{itemize}

\subsection{Snapshots}
Of central importance to the presented dataset is the concept of a snapshot.
A snapshot is herein a set of data gathered together, at a contiguous time
unit.
Examples: the first 20 thousand email messages of an email list
comprises a snapshot; the tweets from the MAMA music event is a
snapshot; the friendship, interaction and posts structures of a facebook
group, prospected at the same time, is a snapshot.
Table~\ref{tsnap} holds the number of snapshots from each provenance in LOSD.

@@

\subsection{Facebook data}
Friendship ego networks (networks whose nodes represent friends of a user and links represent friendships)
were donated from individual users in 2013 and 2014.
Friendship and interaction networks from groups were gathered from
groups where the first author was a participant.
Additionally, some groups have post texts along some metadata, such as
the number of likes.

@@

\subsection{ParticipaBR data}
ParticipaBR is a Brazilian federal platform for social participation,
the most important platform from the myriad of social participation platforms
that were available in 2011-2018~\cite{spbr}.
Texts are derived from blog posts and networks may be obtained from
both friendship and interaction criteria.

@@

\section{Methods}\label{methods}
LOSD is expressed as linked data through RDF
and is ontologicaly described through the
data-driven ontology synthesis method presented in Section~\ref{ont}.

\subsection{Data acquisition}\label{acq}
Research involving human subjects raise well-known ethical issues,
although ethical regulation for (non-medical) research in social systems is not always endorsed by scientists~\cite{eth}.
Furthermore, the online social networking platforms often restrict access
to the data and user may consider illegal the use of their data through web crowlers~\cite{ile1,ile2,ile3}.
Although Twitter and GMANE (email) data is available through public APIs,
the other sources are more problematic.
In this context, research using data, e.g. from Facebook, is precarious
if not prohibited.
We reached a potential solution to this issue through continuated research with
collaborators in the fields of phylosophy, social sciences, antropology and psychology.
The conceptual framework is presented in~\cite{antphy,antphy2}, and in essence
consists in researching first the social networks related to the researchers,
and keeping the procedures, data, algorithms and results publicly available.
These practices enabled collecting and sharing the data in LOSD,
and was called ``Anthropological Physics'' because it relies in the
etnographic tecnique of investigating the social structures
by studying the researcher, as to ammeliorate the ethical issues
(not to be confused e.g. with ``virtual ethnography''~\cite{veth}).
Accordingly, aside from the data acquired from Twitter, all data in LOSD
was gathered from social instances where the first author is a participant.
For example, all Facebook ego networks were donated to the first author by
users which know him and were interested in his research and intervention on the networks,
and all email and Facebook groups data was obtained from groups where the first author is a member.

Facebook data was obtained through the Netvizz software~\cite{netviz},
Facebook allowed downloading friendships and interaction data as available in LOSD.
Data from Twitter was obtained through the Twitter streamming API.
Data From IRC was obtained through bots that kept logs of the messages.
Email data was obtained through the GMANE public API.
Data from ParticipaBR, Cidade Democrática, and AA was obtained through contact
of the researcher with the developers of the platforms.

\subsection{Linked open data}
Linked data refers to data published in the web in such a way that it is
machine readable and conforms to a set of best practices.
The yielded web of data is composed of documents on the web
such as the web of HTML documents.
In practice, the publication of linked data can me summarized
by 1) the use of RDF and 2) the use of RDF
links to interlink data from different sources~\cite{ld1,ld2}.
The web is expected to be interconnected and to
grow by the systematic application of four
steps~\cite{lee1}:
\begin{itemize}
    \item Use URIs to identify things~\cite{uri}.
    \item Use HTTP URIs.
    \item Provide useful information when a URI is accessed via HTTP.
    \item Provide other URIs in the description of resources so human
        and machine agents can perform discovery.
\end{itemize}

@@

\subsection{RDF}
The Resource Description Framework (RDF), a W3C
recommendation, is a model for data
interchange.
It is based on the idea of making statements about resources in the form
of triples, i.e. expressions in the form ``subject - predicate -
object''.
RDF can be serialized in several file formats, including RDF/XML,
Turtle and Manchester, which all, in essence, represent a labeled and
directed multi-graph.
RDF may be stored in a type of database called a triplestore.~\cite{rdf}

As an example of an RDF statement, the following triple in the Turtle
format asserts that ``the paper has color white'':\\
\texttt{http://example.org/Thing\#Paper http://example.org/hasColor\\
http://example.org/Color\#White .}

\subsection{RDFS and OWL ontologies}\label{sont}
One of the most important notions of the semantic web is that of an ontology~\cite{ont}.
An ontology, in this context, is a formalized conceptualization, comprised by
concepts and relations between the concepts and between the relations themselves.
In current semantic web, most simple ontologies are written using the RDFS protocol,
by which one can specify, among other things~\cite{rdfs}:
\begin{itemize}
  \item concepts;
  \item properties, which are concepts that are used as predicates and thus relate concepts;
  \item special relations between concepts that state that one concept is more general than the other\footnote{This
          kind of relation is called hypernymy. Examples: mammal is a hypernym of monkey, drink is a hypernym of beer.};
  \item the subjects and objects that can occurs in a triple where a specific property is a predicate.
\end{itemize}

One can also write an ontology using the OWL protocol,
with which all the expressive capabilities of RDFS are
available, but one can also, among other things~\cite{ont}:
\begin{itemize}
  \item state ``property axioms'', i.e. specify if a property is e.g. reflexive or transitive;
  \item state ``class restrictions'', i.e. specify if a class instance e.g. necessarily holds a relation to another class instance or data.
\end{itemize}
OWL has a richer vocabulary than RDFS and is (way more) complex.
This complexity is a drawback together with the greater computational cost
for performing inference.
The advantage is the greater power to represent conceptualizations.
Using ontologies enables automated reasoning.
For example, if a property \textttt{:r} is known to relate only mokeys to trees,
then, if there is a triple \textttt{:a :r :b}, the instance \textttt{:a} is considered a monkey,
and instance \textttt{:b} is considered a tree,
even if there is no explicit declaration similar
to \textttt{:a rdf:type :Mokey}.

For the data described in this article, the following features are most relevant~\cite{rdfs,ont}:
\begin{itemize}
  \item Using RDFS, one may define a property domain, i.e. the class instances that are allowed as the object in a triple with the property.
  \item Using RDFS, one may define a property range, i.e. the class instances that are allowed as the predicate in a triple with the property.
  \item Using OWL, one may define an existential restriction for a class \textttt{:C},
    i.e. declare that an instance \textttt{:C\#I} of such class has at least one relation with a property \textttt{:p} and a class \textttt{:C2} in the form
    \textttt{:C\#I :p :C2\#I}, where \textttt{:C2\#I} is an instance of the class \textttt{:C2}.
  \item Using OWL, one may define an universal restriction for a class \textttt{:C},
    i.e. declare that any relation of an instance \textttt{:C\#I} of such class, with a property \textttt{:p} of the form \textttt{:C\#I :p :CX\#I},
    implies that \textttt{:CX\#I} is an instance of a specific class, e.g. \textttt{:C2}.
\end{itemize}

\subsection{Data-driven ontology synthesis}\label{ont}
Most ontologies are created by domain experts~\cite{ont}
even though the data they arrange is often given by a software system
and has thus a predefined structure.  
We developed a simple ontology synthesis method that probes
the ontological structure in the data with
SPARQL queries and post-processing.
The results are OWL code and diagrams which are available in the
Supplementary document of this article.
The method can be extended to comprise further OWL axioms and restrictions,
but is currently performed to fit present needs with maximum simplicity.
Present needs are limited to informative figures and
the steps implemented are as follows:
\begin{enumerate}
    \item Obtain all distinct classes with the query:
\begin{lstlisting}[language=spq]
SELECT DISTINCT ?class_uri WHERE { ?s a ?class_uri .}
\end{lstlisting}
\item For each class, obtain the properties that occur as predicates in triples where the subject is an instance of the class:
\begin{lstlisting}[language=spq]
SELECT DISTINCT ?property_uri WHERE { ?s a <class_uri> . ?s ?property_uri ?o .}
\end{lstlisting}
\item Compare the total number of individuals (\te{?cs1}) of the class (\te{class\_uri}) with
	the number of such individuals (\te{?cs2}) that are subjects of at least one triple where 
        the predicate is the property (\te{property\_uri}) and the object is
        an instance of the same class (\te{class2\_uri}).
	The queries are:
\begin{lstlisting}[language=spq]
SELECT (COUNT(DISTINCT ?s) as ?cs1) WHERE { ?s a <class_uri> }
SELECT DISTINCT ?oc WHERE {
  ?s a <class_uri> . ?s <property_uri> ?o . ?o a ?oc .
}
\end{lstlisting}
        Then, for each object class (\textttt{object\_curi}) in \textttt{?oc}:
\begin{lstlisting}[language=spq]
SELECT (COUNT(DISTINCT ?s) as ?cs2) WHERE {
  ?s a <class_uri> . ?s <property_uri> ?o . ?o a <object_curi> .
}
\end{lstlisting}
        When \te{?cs1} matches \te{?cs2}, an existential restriction is found.
        And with the following query, when \te{?cs} is 0, a universal restriction is found:
\begin{lstlisting}[language=spq]
SELECT (COUNT(DISTINCT ?s) as ?cs) WHERE {
  ?s a <class_uri> . ?s <property_uri> ?o . ?o a ?oc .
  FILTER(str(?oc) != 'object_curi')
}
\end{lstlisting}
  \item Retrieve all object classes or datatypes where the subject is an instance of the class and the predicate is the property:
\begin{lstlisting}[language=spq]
SELECT DISTINCT ?co (datatype(?o) as ?do) WHERE {
  ?s a <class_uri>. ?s <property_uri> ?o . OPTIONAL { ?o a ?co . }
}
\end{lstlisting}
  \item Obtain all distinct properties:
\begin{lstlisting}[language=spq]
SELECT DISTINCT ?p WHERE { ?s ?p ?o }
\end{lstlisting}
  \item For each property, check if it is functional, i.e. if it
        occurs at most once with each subject.
        This is performed by counting the objects and further verifying
        that they are at most one. The query is:
\begin{lstlisting}[language=spq]
SELECT DISTINCT (COUNT(?o) as ?co) WHERE {
  ?s <property_uri> ?o 
} GROUP BY ?s
\end{lstlisting}
  \item For each property, find the incident range (possible subjects) and domain (possible objects) with the
        queries:
\begin{lstlisting}[language=spq]
SELECT DISTINCT ?co (datatype(?o) as ?do) WHERE {
  ?s <property_uri> ?o . OPTIONAL { ?o a ?co . }
} 
SELECT DISTINCT ?cs WHERE { ?s <property_uri> ?o . ?s a ?cs . }
\end{lstlisting}
  \item Render diagrams to achieve figures as exemplified in Section~\ref{sdia} and in the Supporting Information document.
\end{enumerate}

\subsection{Standardization}
LOSD data is represented through URIs and triples by translation of various formats to RDF.
URIs are built in the namespace \url{http://purl.org/socialparticipation/participationontology/}
which are identified herein with the prefix ``\te{po:}''.
Classes and properties are built by adding a suffix to the root, as in \te{po:Participant} or \te{po:text}.
Classes have ``UpperCamelCase'' suffixes while properties have ``lowerCamelCase'' suffixes.
All class instances, such as participants, messages, friendships and
interactions, are linked to
snapshots through the triple \te{<instance> po:snapshot <snapshot\_uri>}.
Message texts, including comments, are objects in a triple: \te{<message\_id> po:text '...the text...'}.
Preprocessed texts are objects of triples: \te{<message\_id> po:cleanText '...te text...'}.
More specialized predicates are used for text storage when necessary,
such as \te{po:htmlBodyText} and \te{po:cleanBodyText} used
for ParticipaBR articles.
A participant URI is unique throughout the provenance (e.g. the same for
the same participant in all Twitter snapshots).
To enable annotations which differ when the snapshot changes,
\te{po:Observation} class instances are used in the triple
\te{<participant\_uri> po:observation <observation\_uri>}.
Observation instances are then linked to the snapshot and the
data.

Instance URIs are built using the URI from the class they derive from plus a hashtag character,
a provenance string (e.g. \te{facebook-legacy} or
\textttt{participabr-legacy}), and an identifier;
e.g. \textttt{po:Participant\#<provenance-legacy>-<id>}.
All snapshot URIs follow the formation rule: \textttt{po:<SnapshotProvenance>\#<snapshot\_id>}.
All snapshot ids follow the formation rule: \textttt{<platform>-legacy-<further\_identifier>}; e.g.
\textttt{irc-legacy-labmacambira} or
\textttt{email-legacy-linux.audio.devel1-20000}.

The authors chose to express all URIs in LOSD locally (i.e. within a local namespace), and linkage with external namespaces may be achieved e.g. by using subclass and ``same as'' relations.
Also, each snapshot holds specific \te{po:observation} for each participant, allowing for specific metadata such as provided by the original data or as may be provided uppon analysis.
LOSD does not hold data about each snapshot (such as statistics of the networks), and linkage to external namespaces (such as FOAF and Dublin Core)
to avoid overcomplicating this initial contribution and unecessary use of storage space.

>>>>

%=================================================================
% Full title of the paper (Capitalized)
\Title{LOSD: A Linked Open Social Dataset for Scientific Research}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
The fields of social network analysis and complex networks
are widely researched.
In fact, a myriad of results have been reported which are based in
diverse data, most often not accessible to researchers other than the publishing authors.
In order to enhance the scientific community's common repertoire,
this work presents an open dataset with diverse provenance of 
human social online networking.
Current data was obtained in 2013-16 from well-known online social networking platforms:
Facebook, Twitter, IRC, Email.
Data from Participa.BR, AA, and Cidade Democrática are also supplied as samples of less known instances.
The linked data representation was adopted in order to
comply with current best practices, homogenize access, facilitate discovery
and analyzes which integrate third party and provided data,
and make available a dataset with multiple metadata that admits on-demand retrieval of structures.
Futhermore, the method for obtaing and making the data available is potentially novel,
and may be of use for collecting and sharing additional data and for other research groups.
This document presents a description, usage remarks and overall statistics of the dataset,
which should favor subsequent work in complex and social networks and enhancements to the dataset.
}

@@

\begin{document}
\section{Summary}
The research on human social and complex networks often
relies on data that express real phenomena~\cite{c1,c2,c3}.
Such data is regularly not made publicly available,
which hinders validation of the results
and does not favor subsequent work.
Furthermore, there is a shortcoming of open datasets for benchmarking results
related to complex and social networks,
yielding diverse results from poorly related sources.
Social networks linked datasets are currently most often restricted to a specific platform
or express statistics of social networks, as described in Section~\ref{srel}.
This work presents 
a dataset named LOSD (Linked Open Social Data)
that has been achieved by linking social data derived from diverse
provenance, namely: Facebook, Twitter, IRC, Email groups, Participa.BR, Cidade Democrática, and AA.
Such data may be of use as a common and public repertoire for scientific
research involving complex networks, specially if social networks and textual content are pinpointed.
In fact, LOSD has already been used in scientific research~\cite{stab,thesis} (CNPq grant 140860/2013-4),
and should be further exploited e.g. in researching complex networks and data visualization (FAPESP grant 2017/05838-3)
which relies on networks enriched by textual data and on dynamic (or time-evolving) networks.
Data collection was performed by a potentially innovative method which places the researcher
at the core of data collection, and keeps the resources (e.g. code, data, documents) open,
in order to ammeliorate ethical issues involved in studying human social systems.

This paper is organized as follows. Next subsection briefly discusses related work.
Section~\ref{materials} describes the data currently available in LOSD.
Section~\ref{methods} describes the methods used for data gathering,
translation into RDF linked data, and the synthesis of auxiliary structures.
Section~\ref{usage} holds directions for LOSD usage.

\subsection{Related work}\label{srel}
Most available datasets on human social networks rely only on one source (e.g.~\cite{nat,fb1}),
on specialized communities such as related to research or movies (e.g.~\cite{s1}),
or present only statistics about social networks, and not the data that yields the networks (e.g.~\cite{st1}.
In~\cite{foaf1}, the authors presented a set of tools to obtain linked social data available in the public data cloud,
and made available a dataset from the Advogato platform,
dedicated to free software developers connected by trust relatioships.
There are various datasets which provide data obtained from Twitter (e.g.~\cite{tw1,tw2}),
most often with a specific focus (users, retweets, hashtags, user mentions, etc)
and not presented to the scientific community by means of a research article.
The datasets which are most closely related to LOSD are: KONECT~\cite{konect},
SNAP Datasets~\cite{snapnets}, the Network Repository~\cite{nr}, and ICON~\cite{icon}.
None of them are (RDF) linked data but express the networks in formats specialized for networks.
In total, these datasets make available thousands of networks,
a fraction of them are social networks from diverse provenance and
of various types (e.g. simple, directed, weighted, bipartite, temporal networks).
Compared to LOSD, KONECT, SNAP, Network Repository, and ICON are complementary, providing a further diversity of networks,
but only as simplified data, i.e. in specialized formats, most often expressed as columns of node ids
that are connected and metadata (such as timestaps for temporal networks).
In summary, no linked human social network dataset presented to the
scientific community containts the diverse provenance used in LOSD,
or provides such a wide range of metadata,
potentially because researchers did not have the means to acquire and
share the data as described in Section~\ref{acq},
and because linked data standards are (still) not widely adopted in 
every research domain.

\section{Data Description}\label{materials}
\subsection{Data outline}\label{outline}
% scripts/percolationUse2.py
The LOSD dataset consists of 31,996,740 triples, 350,284 participants,
3,030,433 network links yield by most important interactions or relations, and 245,194,377 characters.
Among all snapshots, 63 are ego snapshots, 54 are group snapshots; 50 have interaction links,
89 have friendship links; 43 have text content associated to participant activity.
Section~\ref{queries} holds the queries used to obtain such data.
This section describes the data in LOSD in very brief terms.
The Supplementary document is provided for further information about the dataset.

\noindent Data was gathered from:
\begin{itemize}
    \item public APIs (Twitter, Email);
    \item public logs (IRC and AA);
    \item Netvizz software~\cite{netvizz} and subsequent donation by users (Facebook);
    \item donation by system administrators (AA, Participa.BR, Cidade Democr\'atica).
\end{itemize}

\subsection{Snapshots}
Of central importance to the presented dataset is the concept of a snapshot.
A snapshot is herein a set of data gathered together, at a contiguous time
unit.
Examples: the first 20 thousand email messages of an email list
comprises a snapshot; the tweets from the MAMA music event is a
snapshot; the friendship, interaction and posts structures of a facebook
group, prospected at the same time, is a snapshot.
Table~\ref{tsnap} denotes the number of snapshots from each provenance in LOSD.

@@

\subsection{Facebook data}
Friendship ego networks (networks whose nodes represent friends of a user and links represent friendships)
were donated from individual users in 2013-5.
Friendship and interaction networks from groups were gathered from
groups where the first author was a participant.
Additionally, some groups have post texts along some metadata, such as
the number of likes.

@@

\subsection{Participa.BR data}
Participa.BR is a Brazilian federal platform for social participation,
one of the most important platforms from the myriad of social participation platforms
that were available in Brazil in 2011-2018~\cite{spbr}.
Texts are derived from blog posts and networks may be obtained from
both friendship and interaction criteria.

@@

\section{Methods}\label{methods}
LOSD is expressed as RDF linked data
and is ontologicaly described through the
data-driven ontology synthesis method presented in Section~\ref{ont}.

\subsection{Data acquisition}\label{acq}
Research involving human subjects raises well-known ethical issues,
although ethical regulation for (non-medical) research in social systems
is not always endorsed by scientists~\cite{eth}.
Furthermore, the online social networking platforms often restrict access
to the data and users may consider illegal the use of their data through web crowlers~\cite{ile1,ile2,ile3}.
Although Twitter and GMANE (email) data are available through public APIs,
the other sources are more problematic.
In this context, research using data, e.g. from Facebook, is precarious
if not prohibited.
We reached a potential solution to this issue through continuated research with
collaborators in the fields of phylosophy, social sciences, antropology and psychology.
The conceptual framework is presented in~\cite{antphy,antphy2,antphy3}, and in essence
consists in researching first the social networks related to the researchers,
and keeping the procedures, data, algorithms and results publicly available.
These practices enabled collecting and sharing the data in LOSD,
and was called ``Anthropological Physics'' because it relies in the
etnographic tecnique of investigating social structures
by studying the researcher, as to ammeliorate the ethical issues
(not to be confused e.g. with ``virtual ethnography''~\cite{veth}).
Accordingly, aside from the data acquired from Twitter, all data in LOSD
was gathered from social instances where the first author was a participant.
For example, all Facebook ego networks were donated to the first author by
users which knew him and were interested in his research and intervention on the networks,
and all email and Facebook groups data was obtained from groups where the first author was a member.

Facebook data was obtained through the Netvizz software~\cite{netviz},
when Facebook allowed downloading friendships and interaction data as available in LOSD.
Data from Twitter was obtained through the Twitter streamming API.
Data From IRC was obtained through bots that kept logs of the messages.
Email data was obtained through the GMANE public API.
Data from Participa.BR, Cidade Democrática, and AA was obtained through contact
of the researcher with the developers of the platforms.
These sources are expreesed in Table~\ref{tab:provenance}.

\subsection{Linked open data}
Linked data refers to data published in the web in such a way that it is
machine readable and conforms to a set of best practices.
The yielded web of data is composed of documents on the web
in similarity to the web of HTML documents.
In practice, the publication of linked data can me summarized
by 1) the use of RDF and 2) the use of RDF
links to interlink data from different sources~\cite{ld1,ld2}.
The web is expected to be interconnected and to
grow by the systematic application of four
steps~\cite{lee1}:
\begin{itemize}
  \item Use URIs (Uniform Resource Identifiers) to identify things~\cite{uri}.
    \item Use HTTP URIs.
    \item Provide useful information when a URI is accessed via HTTP.
    \item Provide other URIs in the description of resources so human
        and machine agents can perform discovery.
\end{itemize}

@@

\subsection{RDF}
The Resource Description Framework (RDF), a W3C
recommendation, is a model for data
interchange.
It is based on the idea of making statements about resources in the form
of triples, i.e. expressions in the form ``subject - predicate -
object''.
RDF can be serialized in several file formats, including RDF/XML,
Turtle and Manchester, which all, in essence, represent a labeled and
directed multigraph.
RDF may be stored in a type of database called a triplestore.~\cite{rdf}

As an example of an RDF statement, the following triple in the Turtle
format asserts that ``the paper has color white'':\\
\te{http://example.org/Thing\#Paper http://example.org/hasColor\\
http://example.org/Color\#White .}

\subsection{RDFS and OWL ontologies}\label{sont}
One of the most important notions of the semantic web is that of an ontology~\cite{ont}.
An ontology, in this context, is a formalized conceptualization, comprised by
concepts and relations between the concepts and between the relations themselves.
In current semantic web, most simple ontologies are written using the RDFS protocol,
by which one can specify, among other things~\cite{rdfs}:
\begin{itemize}
  \item concepts;
  \item properties, which are concepts that are used as predicates and thus relate concepts;
  \item special relations between concepts that state that one concept (superclass) is more general than the other (subclass)\footnote{This
          kind of relation is called a hypernymy. Examples: mammal is a hypernym of monkey, drink is a hypernym of beer.};
  \item the subjects and objects that can occurs in a triple where a specific property is a predicate.
\end{itemize}

One may also write an ontology using the OWL protocol,
with which all the expressive capabilities of RDFS are
available, but one can also, among other things~\cite{ont}:
\begin{itemize}
  \item state ``property axioms'', i.e. specify if a property is e.g. reflexive or transitive;
  \item state ``class restrictions'', i.e. specify if a class instance e.g. necessarily holds a relation to another class instance or data.
\end{itemize}
OWL has a richer vocabulary than RDFS and is (way more) complex.
This complexity is a drawback together with the greater computational cost
for performing inference.
The advantage is the greater power to represent conceptualizations.
Using ontologies enables automated reasoning (usually called ``inference'').
A very simple example: if a property \te{:r} is known to relate only mokeys to trees,
then, if there is a triple \te{:a :r :b}, the instance \te{:a} is considered a monkey,
and instance \te{:b} is considered a tree,
even if there is no explicit declaration similar
to \te{:a rdf:type :Mokey}.

For the data described in this article, the following features are most relevant~\cite{rdfs,ont}:
\begin{itemize}
  \item Using RDFS, one may define a property domain, i.e. the class instances that are allowed as the subject in a triple with the property.
  \item Using RDFS, one may define a property range, i.e. the class instances that are allowed as the object in a triple with the property.
  \item Using OWL, one may define an existential restriction for a class \te{:C} using a property \te{:p} and another class \te{:C2},
    i.e. declare that $\forall$ \te{:C\#I}, if \te{:C\#I rdf:type :C} then $\exists$ \te{:C2\#I rdf:type :C2} such that \te{:C\#I :p :C2\#I}.
  \item Using OWL, one may define a universal restriction for a class \te{:C} using a property \te{:p} and another class \te{:C2},
    i.e. declare that $\forall$ \te{:C\#I}, if \te{:C\#I rdf:type :C} then
    \te{:C\#I :p :C2\#I} implies that \te{:C2\#I rdf:type :C2}.
\end{itemize}

\subsection{Data-driven ontology synthesis}\label{ont}
Most ontologies are created by domain experts~\cite{ont}
even though the data they arrange is often output by a software system
and has thus a predefined structure.  
We developed a simple ontology synthesis method that probes
the ontological structure in the data with
SPARQL queries and post-processing.
The results are OWL code and diagrams which are available in the
Supplementary document of this article.
The method can be extended to comprise further OWL axioms and restrictions,
but is currently performed to fit present needs with maximum simplicity.
Present needs are limited to informative figures and
the steps implemented are as follows:
\begin{enumerate}
    \item Obtain all distinct classes with the query:
\begin{lstlisting}[language=spq]
SELECT DISTINCT ?class_uri WHERE { ?s a ?class_uri .}
\end{lstlisting}
\item For each class, obtain the properties that occur as predicates in triples where the subject is an instance of the class:
\begin{lstlisting}[language=spq]
SELECT DISTINCT ?property_uri WHERE { ?s a <class_uri> . ?s ?property_uri ?o .}
\end{lstlisting}
\item Compare the total number of individuals (\te{?cs1}) of the class (\te{class\_uri}) with
	the number of such individuals (\te{?cs2}) that are subjects of at least one triple where 
        the predicate is the property (\te{property\_uri}) and the object is
        an instance of the same class (\te{object\_curi}).
	The queries are:
\begin{lstlisting}[language=spq]
SELECT (COUNT(DISTINCT ?s) as ?cs1) WHERE { ?s a <class_uri> }
SELECT DISTINCT ?oc WHERE {
  ?s a <class_uri> . ?s <property_uri> ?o . ?o a ?oc .
}
\end{lstlisting}
        Then, for each object class (\te{object\_curi}) in \te{?oc}:
\begin{lstlisting}[language=spq]
SELECT (COUNT(DISTINCT ?s) as ?cs2) WHERE {
  ?s a <class_uri> . ?s <property_uri> ?o . ?o a <object_curi> .
}
\end{lstlisting}
        When \te{?cs1} matches \te{?cs2}, an existential restriction is found.
        And with the following query, when \te{?cs} is 0, a universal restriction is found:
\begin{lstlisting}[language=spq]
SELECT (COUNT(DISTINCT ?s) as ?cs) WHERE {
  ?s a <class_uri> . ?s <property_uri> ?o . ?o a ?oc .
  FILTER(str(?oc) != 'object_curi')
}
\end{lstlisting}
  \item Retrieve all object classes or datatypes where the subject is an instance of the class and the predicate is the property:
\begin{lstlisting}[language=spq]
SELECT DISTINCT ?co (datatype(?o) as ?do) WHERE {
  ?s a <class_uri>. ?s <property_uri> ?o . OPTIONAL { ?o a ?co . }
}
\end{lstlisting}
  \item Obtain all distinct properties:
\begin{lstlisting}[language=spq]
SELECT DISTINCT ?p WHERE { ?s ?p ?o }
\end{lstlisting}
  \item For each property, check if it is functional, i.e. if it
        occurs at most once with each subject.
        This is performed by counting the objects and further verifying
        that they are at most one. The query is:
\begin{lstlisting}[language=spq]
SELECT DISTINCT (COUNT(?o) as ?co) WHERE {
  ?s <property_uri> ?o 
} GROUP BY ?s
\end{lstlisting}
  \item For each property, find the incident range (possible subjects) and domain (possible objects) with the
        queries:
\begin{lstlisting}[language=spq]
SELECT DISTINCT ?co (datatype(?o) as ?do) WHERE {
  ?s <property_uri> ?o . OPTIONAL { ?o a ?co . }
} 
SELECT DISTINCT ?cs WHERE { ?s <property_uri> ?o . ?s a ?cs . }
\end{lstlisting}
  \item Render diagrams to achieve figures as exemplified in Section~\ref{sdia} and provided in the Supplementary document.
\end{enumerate}

\subsection{Standardization}
LOSD data is represented through URIs and triples by translation of various formats to RDF.
URIs are built in the namespace \te{http://purl.org/socialparticipation/participationontology/}
which are identified herein with the prefix ``\te{po:}''.
Classes and properties are built by adding a suffix to the root, as in \te{po:Participant} or \te{po:text}.
Classes have ``UpperCamelCase'' suffixes while properties have ``lowerCamelCase'' suffixes.
All class instances, such as participants, messages, friendships and
interactions, are linked to
snapshots through the triple \te{<instance> po:snapshot <snapshot\_uri>}.
Message texts, including comments, are objects in a triple: \te{<message\_id> po:text '...the text...'}.
Preprocessed texts are objects of triples: \te{<message\_id> po:cleanText '...the text...'}.
More specialized predicates are used for text storage when necessary,
such as \te{po:htmlBodyText} and \te{po:cleanBodyText} used
for Participa.BR articles.
A participant URI is unique throughout the provenance (e.g. the same for
the same participant in all Twitter snapshots).
To enable annotations which differ when the snapshot changes,
\te{po:Observation} class instances are used in the triple
\te{<participant\_uri> po:observation <observation\_uri>}.
Observation instances are then linked to the snapshot and the
data.

Instance URIs are built using the URI from the class they derive from plus a hashtag character,
a provenance string (e.g. \te{facebook-legacy} or
\te{participabr-legacy}), and an identifier;
e.g. \te{po:Participant\#<provenance-legacy>-<id>}.
All snapshot URIs follow the formation rule: \te{po:<SnapshotProvenance>\#<snapshot\_id>}.
All snapshot ids follow the formation rule: \te{<platform>-legacy-<further\_identifier>}; e.g.
\te{irc-legacy-labmacambira} or
\te{email-legacy-linux.audio.devel1-20000}.

The authors chose to express all URIs in LOSD locally (i.e. within a local namespace), and linkage with external namespaces may be achieved e.g. by using subclass and ``same as'' relations.
Also, each snapshot holds a specific \te{po:Observation} instance related to each participant,
allowing for specific metadata such as provided by the original data or as may be provided uppon analysis.
Current LOSD does not hold statistics about each snapshot, or linkage to external namespaces (such as FOAF and Dublin Core),
to avoid overcomplicating this initial contribution and unecessary use of storage space.
